Shaping strategies and standardisation of faceted spheroids from Qianshangying (North China)

### 2025-06-25 analysis/paper/paper.qmd
comments and edits to intro and site context

The pursuit of perfect geometric forms—particularly the ideal sphere—appears to transcend time, persisting from antiquity to the present. Spheroids may be kind of stone artifacts that embody this aspiration toward geometric perfection, reflecting Paleolithic humans’ spatial and navigational cognition, and their grasp of abstract concepts such as ''3D symmetry'' ([@Muller2023; @clark1955; @wynn2002]). This cognitive awareness may have emerged even earlier than the production of handaxes ([@beyene2013; @shipton2019; @white2018]). Consequently, the study of spheroids—including their manufacturing techniques and morphological characteristics—has become a key issue in interpreting early human technological behaviors and cognitive capacities ([@cabanès2022]).

@@

Although these artifact types are frequently grouped under the category “PSB” (or “PSSB” when subspheroids are included) and are commonly interpreted as volumetrically reduced artifacts organized around a central point, this morphology-based classification has led to several interpretive challenges. Artifacts identified as spheroids in one study may be classified as bolas or attributed to entirely different technological categories in another. Furthermore, research has shown that these forms may not represent a unified *chaîne opératoire*. Differences in raw material selection suggest divergent technological strategies and potentially distinct functional purposes ([@deweyer2017; @jones1994]).

At present, the technological category and function of spheroids remain central topics of debate. Broadly, spheroids have been interpreted as hammerstones ([@mussi2025; @schick1994], cores ([@sahnouni1997]), or projectiles ([@leakey1979]). Closely tied to this debate is the question of their production strategy, which remains unclear. Experimental replication and site-specific diacritical analyses have produced divergent interpretations. One view suggests that spheroids emerged unintentionally as the by-products of core exhaustion, which represents a low-effort strategy aimed at flake production ([@sahnouni1997]). Another perspective holds that spheroids were naturally produced through high-intensity battering ([@schick1994]; \[\@mora2005\]; [@clark1955]). In contrast, a third view proposes that spheroids reflect a conceptual template and were produced through a deliberate shaping strategy ([@tittonSubspheroidsLithicAssemblage2020; @Muller2023; @texier2014]).

Spheroids are a widely distributed type of stone artifact, appearing across a broad temporal and geographic range—from the Oldowan and Acheulean to the Middle Paleolithic—and in diverse archaeological assemblages ([@willoughby1985; @Cabanes2024]). This wide distribution is matched by considerable technological and morphological variability, which has fueled ongoing debates concerning their methods of production and functional significance. Site-specific studies are therefore essential for reconstructing the spatial and temporal distribution of spheroids, as well as for identifying their distinct technological traits and functions. Such analyses provide a foundation for situating spheroids within broader discussions of early human technological behavior.

In this study, we analyzed material from the Middle Pleistocene Qianshangying site (QSY) in northern China. Among an assemblage previously classified as a simple core-flake industry, we identified a group of highly standardized spheroids and subspheroids. we observed notable differences between these artifacts and multifacial cores or polyhedrons, which are stone artifacts through typical flaking strategies. We proposed that their concept and manufacture reflect a distinct technological tradition, separate from conventional flake-based technology.

We conducted quantitative morphological and technological analyses, including the novel spherical harmonic analysis, on the spheroids recovered from the QSY site. Our goal was to evaluate whether these spheroids exhibited standardized morphology and production patterns, and whether their tendency toward spherical form resulted from deliberate and systematic reduction processes. These analyses aimed to clarify the technological strategy behind spheroid production at QSY, within the broader context of Middle Pleistocene lithic practices in northern China.

@@

The QSY lies in the Nihewan Basin, a region in northern China situated between the Chinese Loess Plateau and the Inner Mongolian Plateau. The basin preserves extensive Quaternary fluvio-lacustrine and loess deposits ([@li2000; @yuan2009]), hosts the highest concentrations of Pleistocene Paleolithic sites in East Asia, and is especially significant for preserving evidence of early hominin activities outside Africa dating back over one million years ([@yang2020; @yang2017; @pei2019; @pei2017; @zhu2001; @zhu2004]). In recent years, nearly 20 Middle Pleistocene sites—including Jijiazhuang and Caijiagou—have been identified in the southeastern part of the basin ([@yuwei2023; @ye2024]), within the Yuxian Sub-basin, thereby completing the regional cultural sequence of the Nihewan Basin.

Qianshangying site complex consisting of 4 localities from QSY-A to QSY-D, discovered in 2015, is among the region’s most artifact-rich archaeological sites. Geomorphological and sedimentological evidence indicates that the stratigraphy at the site records a full cycle of lake expansion, retreat, re-expansion, and eventually disappearance of the ancient Nihewan lake, and hominin occupation occurred in a marginal lacustrine setting during a low lake-level episode.

The stratigraphic context of QSY indicates a short-term occupation resulting from a single hominin activity event. The assemblage appears to be largely in situ, exhibiting minimal disturbance after deposition and well-preserved patterns of technological combination. ESR dating of the cultural layer places hominin activity at approximately 429 ± 39 ka.

A total of 957 stone artifacts were recovered from four localities at the QSY site, with QSY-A and QSY-B yielding the most abundant assemblages. Following a widely used techno-typological framework (Isaac, 1981, 1986; de la Torre and Mora, 2005, 2018; de la Torre et al., 2011; Pei et al., 2017), the assemblage was classified into four general categories: flaked pieces (n = 247, 25.9%), detached pieces (n = 686, 71.9%), pounded pieces (n = 4, 0.4%), and unmodified materials (n = 17, 1.8%).

>>>>

The pursuit of perfect geometric forms—particularly the ideal sphere—appears to transcend time, persisting from antiquity to the present. Spheroids may be kind of stone artifacts that embody this aspiration toward geometric perfection, reflecting Paleolithic humans’ spatial and navigational cognition, and their grasp of abstract concepts such as ''3D symmetry'' ([@Muller2023; @clark1955; @wynn2002]). This cognitive awareness may have emerged even earlier than the production of handaxes ([@beyene2013; @shipton2019; @white2018]). Consequently, the study of spheroids—including their manufacturing techniques and morphological characteristics—has become a key issue in interpreting early human technological behaviors and cognitive capacities ([@cabanès2022]). [BM: this paragraph seems too poetic and abstract for a scholarly journal article. I think it will be better to open the paper with your sentence below that starts with 'Spheroids are a widely distributed type...']

@@

Although these artifact types are frequently grouped under the category “PSB” (or “PSSB” when subspheroids are included) [BM: we should fully spell out the abbreviation at the first time we show it to the reader] and are commonly interpreted as volumetrically reduced artifacts organized around a central point, this morphology-based classification has led to several interpretive challenges. Artifacts identified as spheroids in one study may be classified as bolas or attributed to entirely different technological categories in another. Furthermore, research has shown that these forms may not represent a unified *chaîne opératoire* [BM: if we say 'research has shown', then we need to cite that research in the same sentence]. Differences in raw material selection suggest divergent technological strategies and potentially distinct functional purposes ([@deweyer2017; @jones1994]).

At present, the technological category and function of spheroids remain central topics of debate. Broadly, spheroids have been interpreted as hammerstones ([@mussi2025; @schick1994], cores ([@sahnouni1997]), or projectiles ([@leakey1979]). Closely tied to this debate is the question of their production strategy, which remains unclear. Experimental replication and site-specific diacritical analyses have produced divergent interpretations. One view suggests that spheroids emerged unintentionally as the by-products of core exhaustion, which represents a low-effort strategy aimed at flake production ([@sahnouni1997]). Another perspective holds that spheroids were naturally produced through high-intensity battering ([@schick1994]; \[\@mora2005\]; [@clark1955]). In contrast, a third view proposes that spheroids reflect a conceptual template and were produced through a deliberate shaping strategy ([@tittonSubspheroidsLithicAssemblage2020; @Muller2023; @texier2014]). [BM: yes, this is great, good job, this is a very important paragraph for the introduction of the paper. We might need to expand it a bit]

Spheroids are a widely distributed type of stone artifact, appearing across a broad temporal and geographic range—from the Oldowan and Acheulean to the Middle Paleolithic—and in diverse archaeological assemblages ([@willoughby1985; @Cabanes2024]). This wide distribution is matched by considerable technological and morphological variability, which has fueled ongoing debates concerning their methods of production and functional significance. [BM: you have already mentioned debates in the previous paragraph, so it's seems a bit odd to mention debates again here without continuity with the previous paragraph. Maybe these two sentences are better as the opening sentences for the paper, setting the scene for more information about the debates] Site-specific studies are therefore essential for reconstructing the spatial and temporal distribution of spheroids, as well as for identifying their distinct technological traits and functions. Such analyses provide a foundation for situating spheroids within broader discussions of early human technological behavior.

In this study, we analyzed material from the Middle Pleistocene Qianshangying site (QSY) in northern China. Among an assemblage previously classified as a simple core-flake industry, we identified a group of highly standardized spheroids and subspheroids. we observed notable differences between these artifacts and multifacial cores or polyhedrons, which are stone artifacts through typical flaking strategies. We proposed that their concept and manufacture reflect a distinct technological tradition, separate from conventional flake-based technology. [BM: I think it would be good to rephrase this as a question that the rest of the paper will answer, that will help to create a sense of flow for the reader.]

We conducted quantitative morphological and technological analyses, including the novel spherical harmonic analysis, on the spheroids recovered from the QSY site [BM: perhaps something like 'a novel application of spherical harmonic analysis methods to an East Asian palaeolithic assemblage']. Our goal was to evaluate whether these spheroids exhibited standardized morphology and production patterns, and whether their tendency toward spherical form resulted from deliberate and systematic reduction processes [BM: this is a very nice and clear statement of our goal, I think it might be good to put it at the second sentence of the paragraph before this one, so we state our goal, then we say how we will tackle it]. These analyses aimed to clarify the technological strategy behind spheroid production at QSY, within the broader context of Middle Pleistocene lithic practices in northern China.

@@

QSY lies in the Nihewan Basin, a region in northern China situated between the Chinese Loess Plateau and the Inner Mongolian Plateau. The basin preserves extensive Quaternary fluvio-lacustrine and loess deposits ([@li2000; @yuan2009]), hosts the highest concentrations of Pleistocene Paleolithic sites in East Asia, and is especially significant for preserving evidence of early hominin activities outside Africa dating back over one million years ([@yang2020; @yang2017; @pei2019; @pei2017; @zhu2001; @zhu2004]). In recent years, nearly 20 Middle Pleistocene sites—including Jijiazhuang and Caijiagou—have been identified in the southeastern part of the basin ([@yuwei2023; @ye2024]), within the Yuxian Sub-basin, thereby completing the regional cultural sequence of the Nihewan Basin. [BM: great detail about the context, can we make a figure that includes a map and a photo of the site?]

The QSY site complex consists of four localities from QSY-A to QSY-D, discovered in 2015. It is among the region’s most artifact-rich archaeological site. Geomorphological and sedimentological evidence indicates that the stratigraphy at the site records a full cycle of lake expansion, retreat, re-expansion, and eventually disappearance of the ancient Nihewan lake, and hominin occupation occurred in a marginal lacustrine setting during a low lake-level episode [BM: citation needed]. The stratigraphic context of QSY indicates a short-term occupation resulting from a single hominin activity event. The assemblage appears to be largely in situ, exhibiting minimal disturbance after deposition and well-preserved patterns of technological combination. ESR dating of the cultural layer places hominin activity at approximately 429 ± 39 ka [BM: citation needed].

A total of 957 stone artifacts were recovered from four localities at the QSY site, with QSY-A and QSY-B yielding the most abundant assemblages. Following a widely used techno-typological framework (Isaac, 1981, 1986; de la Torre and Mora, 2005, 2018; de la Torre et al., 2011; Pei et al., 2017), the assemblage was classified into four general categories: flaked pieces (n = 247, 25.9%), detached pieces (n = 686, 71.9%), pounded pieces (n = 4, 0.4%), and unmodified materials (n = 17, 1.8%). [BM: is it possible to use inline R code for these numbers?]
